---
title: "Analysis walkthrough: hypertension prevalence"
format: html
vignette: >
  %\VignetteIndexEntry{Analysis walkthrough: hypertension prevalence}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r results='hide', message=FALSE, warning=FALSE}
library(chmsflow)
library(recodeflow)
library(dplyr)
```

## Overview

This tutorial walks through a complete analysis using chmsflow: estimating hypertension prevalence across CHMS cycles 3--6. By the end, you will have a harmonized dataset with hypertension status derived for each respondent across four survey cycles.

The steps are:

1. Load cycle data and medication data
2. Recode medications and merge into cycle data
3. Derive hypertension status and its input variables
4. Inspect results
5. Combine cycles into one dataset

## 1. Load data

chmsflow includes bundled dummy data for all six CHMS cycles. Each cycle has a main dataset (`cycle3`, `cycle4`, etc.) and a medication dataset (`cycle3_meds`, `cycle4_meds`, etc.).

```{r}
# Bundled dummy data is loaded automatically with library(chmsflow)
dim(cycle3)
dim(cycle3_meds)
```

At an RDC, you would load real CHMS data from Stata or SAS files instead. See `vignette("using_chmsflow_at_an_rdc")` for setup instructions.

## 2. Recode medications

Medication recoding must happen before deriving hypertension status, because the hypertension derivation uses `any_htn_med` as an input. For full details on medication recoding, see `vignette("recoding_medications")`.

`recode_meds_cycles3to6()` classifies each medication row by drug class, aggregates to one row per respondent, and merges the result into the main cycle data.

```{r, warning=FALSE}
cycle3 <- recode_meds_cycles3to6(cycle3, cycle3_meds, c("any_htn_med", "diab_med"))
cycle4 <- recode_meds_cycles3to6(cycle4, cycle4_meds, c("any_htn_med", "diab_med"))
cycle5 <- recode_meds_cycles3to6(cycle5, cycle5_meds, c("any_htn_med", "diab_med"))
cycle6 <- recode_meds_cycles3to6(cycle6, cycle6_meds, c("any_htn_med", "diab_med"))
```

After this step, `any_htn_med` and `diab_med` are columns in each cycle dataset.

```{r}
select(cycle3, clinicid, any_htn_med, diab_med) |> head()
```

## 3. Derive hypertension status

Use `recode_after_meds()` instead of `rec_with_table()` for variables that depend on medication status. It automatically excludes medication-specific rows from `variable_details` so the pre-computed medication columns are passed through rather than re-derived.

Hypertension derivation requires a chain of intermediate variables: blood pressure measurements, comorbidities (diabetes, CVD, CKD), and their inputs. List all of them so `recode_after_meds()` can derive them in order.

```{r, warning=FALSE}
htn_variables <- c(
  # Blood pressure (raw + adjusted)
  "bpmdpbps", "bpmdpbpd", "sbp_adj_mmhg", "dbp_adj_mmhg",
  # Medication inputs (merged in step 2)
  "any_htn_med", "ccc_32",
  # Diabetes chain (input to hypertension functions)
  "lab_hba1", "diab_a1c", "ccc_51", "diab_med", "diab_status",
  # CVD chain
  "ccc_61", "ccc_63", "ccc_81", "cvd_status",
  # CKD chain
  "lab_bcre", "pgdcgt", "clc_sex", "clc_age", "gfr_ml_min", "ckd_status",
  # Hypertension outcomes
  "htn_status", "htn_adj_status", "htn_control_status", "htn_control_adj_status"
)

cycle3_htn <- recode_after_meds(cycle3, htn_variables)
cycle4_htn <- recode_after_meds(cycle4, htn_variables)
cycle5_htn <- recode_after_meds(cycle5, htn_variables)
cycle6_htn <- recode_after_meds(cycle6, htn_variables)
```

## 4. Inspect results

Each result contains the derived hypertension variables alongside `clinicid`. Values are:

- `1` -- high blood pressure
- `2` -- normal blood pressure
- `tagged_na("a")` -- not applicable
- `tagged_na("b")` -- missing

```{r}
cycle3_htn |>
  select(clinicid, htn_status, htn_adj_status) |>
  head(10)
```

To see the distribution:

```{r}
table(cycle3_htn$htn_status, useNA = "always")
```

For more on how chmsflow handles missing data codes, see `vignette("tagged_na_usage")`.

## 5. Combine cycles

Add a cycle identifier to each dataset and bind them together.

```{r}
cycle3_htn$cycle <- 3
cycle4_htn$cycle <- 4
cycle5_htn$cycle <- 5
cycle6_htn$cycle <- 6

combined <- bind_rows(cycle3_htn, cycle4_htn, cycle5_htn, cycle6_htn)
dim(combined)
```

Hypertension prevalence by cycle:

```{r}
combined |>
  filter(!is.na(htn_status)) |>
  group_by(cycle) |>
  summarise(
    n = n(),
    n_htn = sum(htn_status == 1),
    prevalence = mean(htn_status == 1)
  )
```

## Next steps

- **Medication recoding details** -- `vignette("recoding_medications")`
- **Understanding the metadata** -- `vignette("variables_and_variable_details")`
- **Derived variables** -- `vignette("derived_variables")`
- **Missing data** -- `vignette("tagged_na_usage")`
- **Methodology** -- `vignette("methodology")`
