---
title: "Derived variables"
format: html
vignette: >
  %\VignetteIndexEntry{Derived variables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

There are two types derived variables in the CHMS surveys. Both types of derived variables are supported in `chmsflow`. 

- Variable mapping - mapping two or more variables into a single variable.
- Variables that derived using math equations.

`chmsflow` calculates these more complex derived variables using functions that are referenced in `variable_details.csv` within `RecTo` section with the prefix 'Func::'. The variables used in the function are referenced in the `variableStart` section with the prefix 'DerivedVar::'. For example, GFR (`gfr`) includes `Func::calculate_gfr` in the `RecTo` section; and `DerivedVar::[lab_bcre, pgdcgt, clc_sex, clc_age]` in the `variableStart` section, which indicates the four starting variables (`lab_bcre, pgdcgt, clc_sex, clc_age`).

## Example - Glomerular filtration rate (GFR)

A derived variable for GFR has been created in `chmsflow` that uses harmonized blood creatine (lab_bcre - in µmol/L), ethnicity (pgdcgt - 13 categories), sex (clc_sex - 2 categories), and age (clc_age - in years) variables across all CHMS cycles.

Using `rec_with_table()` you can transform the derived WHR variable across multiple CHMS cycles and create a transformed dataset.

In order derive variables, you must load the existing custom function associated with the derived variable
```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Function for derived GFR variable
calculate_gfr <- function(lab_bcre, pgdcgt, clc_sex, clc_age) {
  GFR <- 0
  serumcreat <- 0

  if (any(!lab_bcre %in% 14:785) || (any(!clc_sex %in% c(1, 2)) || any(!pgdcgt %in% 1:13)) || any(!clc_age %in% 3:79)) {
    GFR <- haven::tagged_na("b") # GFR is NA if any non-responses found
  } else {
    serumcreat <- lab_bcre / 88.4 # Proceeds without non-responses

    if (!is.na(clc_sex) && !is.na(pgdcgt) && serumcreat != 0) {
      if (clc_sex == 2 && pgdcgt == 2) {
        GFR <- 175 * ((serumcreat)^(-1.154)) * ((clc_age)^(-0.203)) * (0.742) * (1.210) # female and black
      } else if (clc_sex == 2 && pgdcgt != 2) {
        GFR <- 175 * ((serumcreat)^(-1.154)) * ((clc_age)^(-0.203)) * (0.742) # female and not black
      } else if (clc_sex == 1 && pgdcgt == 2) {
        GFR <- 175 * ((serumcreat)^(-1.154)) * ((clc_age)^(-0.203)) * (1.210) # male and black
      } else if (clc_sex == 1 && pgdcgt != 2) {
        GFR <- 175 * ((serumcreat)^(-1.154)) * ((clc_age)^(-0.203)) # male and not black
      }
    } else {
      GFR <- haven::tagged_na("b") # Handle case where clc_sex or pgdcgt is NA or serumcreat is 0
    }
  }

  return(GFR)
}
```

```{r, warning=FALSE, message=FALSE}
library(chmsflow)
```

```{r, warning = FALSE, eval=FALSE}
cycle2_gfr <- recodeflow:::rec_with_table(cycle2, variables = c("gfr", "lab_bcre", "pgdcgt", "clc_sex", "clc_age"), variable_details = variable_details, log = TRUE)
```

Since derived variables are based on previously transformed variables, if you want to only transform your derived variable, you must also specify its base CHMS variables in `rec_with_table()` as shown above. So for the derived GFR variable, you will have to also specify the blood creatine (lab_bcre), ethnicity (pgdcgt), sex (clc_sex), and age (clc_age) variables.

## Creating a derived variable

Creating a derived variable requires the harmonization of existing CHMS variables, and a custom function that uses those harmonized variables. For more information on how to create a derived variable [see here](how_to_add_variables.html)
