---
title: "Recoding CHMS variables - Getting Started!"
format: html
---

1. Install chmsflow locally.
```{r}
devtools::document()
devtools::install()
```

2. Load worksheets (variables and variable details sheets).
```{r}
library(here) # Ensure here is set to your current working directory

variables <- read.csv(here("inst", "extdata", "variables.csv"))
variable_details <- read.csv(here("inst", "extdata", "variable-details.csv"))
```

3. Create CHMS synthetic data with missing values. Cycle data object always has to be called "cyclex" in order for recoding to work.

At the RDC, each cycle is split into multiple components (e.g., household data, clinic data, laboratory data, etc.), so it is the analyst's responsibility to merge their required components in one database named "cyclex". However, keep medication data separate from the rest of the cycle data (see recoding-medications.qmd). 

Note that row headers for cycle 6 must be put to lower case prior to recoding.
```{r}
# Create synthetic data with missing values
n <- 100  # Number of observations
set.seed(123)  # Set seed for reproducibility

# Generate random ages between 18 and 75, with some missing values [996-999]
ages <- sample(c(18:75), n, replace = TRUE)

# Generate random sexes (1 for male, 2 for female, with some missing values [6-9])
sexes <- sample(c(1:2), n, replace = TRUE)

# Generate random systolic blood pressures between 80 and 215, with some missing values [996-999]
systolic_bps <- sample(c(80:215, 996:999), n, replace = TRUE)

# Generate random diastolic blood pressures between 50 and 150, with some missing values [996-999]
diastolic_bps <- sample(c(50:150, 996:999), n, replace = TRUE)

# Create a data frame named cycle4
cycle4 <- data.frame(clc_age = ages, clc_sex = sexes, bpmdpbps = systolic_bps, bpmdpbpd = diastolic_bps)

# Print the first few rows of the cycle4 data
print(head(cycle4))
```

4. Test recodeflow's rec_with_table on fake CHMS data.
```{r}
# Recoding basic variables
cycle4_ages <- recodeflow::rec_with_table(cycle4, "clc_age", variable_details = variable_details, log = TRUE)
head(cycle4_ages)

cycle4_sexes <- recodeflow::rec_with_table(cycle4, "clc_sex", variable_details = variable_details, log = TRUE)
head(cycle4_sexes)

# Recoding a transformed variable (cont to cat)
cycle4_categorical_ages <- recodeflow::rec_with_table(cycle4, "agegroup4", variable_details = variable_details, log = TRUE)
head(cycle4_categorical_ages)

# Recoding a derived variable
cycle4_adjusted_SBPs <- recodeflow::rec_with_table(cycle4, c("bpmdpbps", "sbp_adj"), variable_details = variable_details, log = TRUE)
head(cycle4_adjusted_SBPs)
```