---
title: "Recoding CHMS medication variables"
format: html
execute:
  eval: false
vignette: >
  %\VignetteIndexEntry{Recoding CHMS medication variables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r results='hide', message=FALSE, warning=FALSE, eval=TRUE}
library(chmsflow)
```

## 1. Introduction

chmsflow provides 16 functions that classify medications from ATC codes recorded in CHMS clinic data. Each function checks whether a respondent is taking a specific drug class and returns `1` (yes) or `0` (no), with `haven::tagged_na()` codes for missing or not-applicable responses.

### Available medication variables

| Variable | Drug class | ATC prefix | Cycles 3-6 function | Cycles 1-2 function |
|----------|------------|------------|----------------------|----------------------|
| `ace_med` | ACE inhibitors | C09 | `is_ace_inhibitor()` | `is_ace_med_cycles1to2()` |
| `bb_med` | Beta blockers | C07 | `is_beta_blocker()` | `is_bb_med_cycles1to2()` |
| `ccb_med` | Calcium channel blockers | C08 | `is_calcium_channel_blocker()` | `is_ccb_med_cycles1to2()` |
| `diur_med` | Diuretics | C03 | `is_diuretic()` | `is_diur_med_cycles1to2()` |
| `misc_htn_med` | Other antihypertensives | mixed | `is_other_antihtn_med()` | `is_misc_htn_med_cycles1to2()` |
| `any_htn_med` | Any antihypertensive | combined | `is_any_antihtn_med()` | `is_any_htn_med_cycles1to2()` |
| `nsaid_med` | NSAIDs | M01A | `is_nsaid()` | `is_nsaid_med_cycles1to2()` |
| `diab_med` | Diabetes medications | A10 | `is_diabetes_med()` | `is_diab_med_cycles1to2()` |

### Cycle differences

Medication data is structured differently across CHMS cycles:

- **Cycles 1-2** store medications in a flat format with up to 80 individual columns (`atc_101a` to `atc_235a` for ATC codes, `mhr_101b` to `mhr_235b` for time last taken). The cycles 1-2 wrapper functions accept all of these columns as parameters.
- **Cycles 3-6** store medications in a multi-row format with two variables per row: `meucatc` (ATC code) and `npi_25b` (time last taken). Each respondent may have multiple rows. After recoding, results must be aggregated by `clinicid`.

## 2. When to use medication recoding

If your analysis requires medication variables, ***always perform medication recoding first***, before recoding any other variables.

Two downstream health outcome variables depend on medication status:

- **Hypertension** - `any_htn_med` is a required input for `derive_hypertension()`, `derive_hypertension_adj()`, `derive_hypertension_control()`, and `derive_hypertension_control_adj()`.
- **Diabetes** - `diab_med` is a required input for `derive_diabetes_status()`.

Use `recode_meds_cycles1to2()` or `recode_meds_cycles3to6()` to derive medication variables and merge them into the main cycle dataset before recoding health outcomes. This avoids passing long lists of raw medication columns into downstream `rec_with_table()` calls.

## 3. Quick start

A minimal example for cycles 3-6: recode ACE inhibitor status, aggregate per respondent, and merge with main cycle data.

```{r, warning=FALSE}
library(recodeflow)
library(dplyr)

# Recode ace_med from multi-row medication data
cycle3_ace_data <- rec_with_table(
  cycle3_meds,
  c("clinicid", "meucatc", "npi_25b", "ace_med"),
  variable_details = variable_details
)

# Aggregate to one row per respondent (take max: 1 if any row matched)
cycle3_ace_data <- cycle3_ace_data |>
  group_by(clinicid) |>
  summarize(ace_med = max(as.numeric(as.character(ace_med))))

# Merge with main cycle data
cycle3 <- merge(cycle3, cycle3_ace_data, by = "clinicid")
head(select(cycle3, clinicid, ace_med))
```

## 4. Detailed workflow

### 4.1 Recode medication variables for individual cycles

The medication data object must always be named `cyclex_meds` for the recoding scripts to function correctly. All other data for the same cycle should be stored separately under the name `cyclex`, as specified in `recoding-variables.qmd`.

```{r, warning=FALSE}
library(recodeflow)
library(dplyr)

# Recoding basic variables from cycles 2 and 3
cycle2_medication_data <- rec_with_table(
  cycle2_meds,
  c("clinicid", "atc_101a", "mhr_101b"),
  variable_details = variable_details
)
head(cycle2_medication_data)

cycle3_medication_data <- rec_with_table(
  cycle3_meds,
  c("clinicid", "meucatc", "npi_25b"),
  variable_details = variable_details
)
head(cycle3_medication_data)
```

#### Cycles 1-2 (flat format)

For cycles 1-2, all raw ATC code and time-last-taken columns must be listed explicitly in the `rec_with_table()` call:

```{r, warning=FALSE}
# Recoding derived variable ace_med for cycle 2
cycle2_ace_medication_data <- rec_with_table(
  cycle2_meds,
  c(
    "clinicid",
    "atc_101a", "atc_102a", "atc_103a", "atc_104a", "atc_105a",
    "atc_106a", "atc_107a", "atc_108a", "atc_109a", "atc_110a",
    "atc_111a", "atc_112a", "atc_113a", "atc_114a", "atc_115a",
    "atc_201a", "atc_202a", "atc_203a", "atc_204a", "atc_205a",
    "atc_206a", "atc_207a", "atc_208a", "atc_209a", "atc_210a",
    "atc_211a", "atc_212a", "atc_213a", "atc_214a", "atc_215a",
    "atc_131a", "atc_132a", "atc_133a", "atc_134a", "atc_135a",
    "atc_231a", "atc_232a", "atc_233a", "atc_234a", "atc_235a",
    "mhr_101b", "mhr_102b", "mhr_103b", "mhr_104b", "mhr_105b",
    "mhr_106b", "mhr_107b", "mhr_108b", "mhr_109b", "mhr_110b",
    "mhr_111b", "mhr_112b", "mhr_113b", "mhr_114b", "mhr_115b",
    "mhr_201b", "mhr_202b", "mhr_203b", "mhr_204b", "mhr_205b",
    "mhr_206b", "mhr_207b", "mhr_208b", "mhr_209b", "mhr_210b",
    "mhr_211b", "mhr_212b", "mhr_213b", "mhr_214b", "mhr_215b",
    "mhr_131b", "mhr_132b", "mhr_133b", "mhr_134b", "mhr_135b",
    "mhr_231b", "mhr_232b", "mhr_233b", "mhr_234b", "mhr_235b",
    "ace_med"
  ),
  variable_details = variable_details
)
head(select(cycle2_ace_medication_data, clinicid, ace_med))
```

#### Cycles 3-6 (multi-row format)

For cycles 3-6, only `meucatc` and `npi_25b` are needed:

```{r, warning=FALSE}
# Recoding derived variable ace_med for cycle 3
cycle3_ace_medication_data <- rec_with_table(
  cycle3_meds,
  c("clinicid", "meucatc", "npi_25b", "ace_med"),
  variable_details = variable_details
)
head(cycle3_ace_medication_data)
```

## 5. Integration with health outcomes

Two downstream variables depend on medication status:

- **Hypertension** — `any_htn_med` is required for `derive_hypertension()` and its variants.
- **Diabetes** — `diab_med` is required for `derive_diabetes_status()`.

The workflow is the same for all cycles: recode medication variables and merge into the main cycle dataset, then derive health outcomes.

Use `recode_after_meds()` instead of `rec_with_table()` when deriving variables that depend on pre-computed medication inputs. It automatically excludes medication-specific rows from `variable_details` so the pre-computed columns are passed through rather than re-derived.

### 5.1 Cycles 3-6

**Step 1** — Recode medication variables and merge with main cycle data. Requires: `cycle3`, `cycle3_meds`.

```{r, warning=FALSE}
library(recodeflow)
library(dplyr)

cycle3 <- cycle3 |>
  left_join(
    recode_meds_cycles3to6(cycle3_meds, c("any_htn_med", "diab_med")),
    by = "clinicid"
  )
```

**Step 2** — Derive diabetes status. Requires: `cycle3` from Step 1.

```{r, warning=FALSE}
cycle3_diab_data <- recode_after_meds(
  cycle3,
  c("lab_hba1", "diab_a1c", "diab_med", "ccc_51", "diab_status")
)
head(select(cycle3_diab_data, clinicid, diab_status))
```

**Step 3** — Derive hypertension status. Requires: `cycle3` from Step 1.

`cvd_status`, `diab_status`, and `ckd_status` are intermediate inputs to the hypertension functions. Their full input chains must also be listed so `recode_after_meds()` can derive them.

```{r, warning=FALSE}
cycle3_htn_data <- recode_after_meds(
  cycle3,
  c(
    # Blood pressure (raw + adjusted)
    "bpmdpbps", "bpmdpbpd", "sbp_adj_mmhg", "dbp_adj_mmhg",
    # Medication inputs (pre-computed in cycle3_with_meds)
    "any_htn_med", "ccc_32",
    # Diabetes chain (input to htn functions)
    "lab_hba1", "diab_a1c", "ccc_51", "diab_med", "diab_status",
    # CVD chain
    "ccc_61", "ccc_63", "ccc_81", "cvd_status",
    # CKD chain
    "lab_bcre", "pgdcgt", "clc_sex", "clc_age", "gfr_ml_min", "ckd_status",
    # Hypertension outcomes
    "htn_status", "htn_adj_status", "htn_control_status", "htn_control_adj_status"
  )
)
head(select(cycle3_htn_data, clinicid, htn_status, htn_adj_status))
```

### 5.2 Cycles 1-2

Cycles 1-2 medication data uses uppercase column names (`CLINICID`, `ATC_101A`, etc.). `recode_meds_cycles1to2()` normalizes these internally.

**Step 1** — Recode medication variables and merge with main cycle data. Requires: `cycle1`, `cycle1_meds`.

```{r, warning=FALSE}
library(recodeflow)
library(dplyr)

cycle1 <- cycle1 |>
  left_join(
    recode_meds_cycles1to2(cycle1_meds, c("any_htn_med", "diab_med")),
    by = "clinicid"
  )
```

**Step 2** — Derive diabetes status. Requires: `cycle1` from Step 1.

```{r, warning=FALSE}
cycle1_diab_data <- recode_after_meds(
  cycle1,
  c("lab_hba1", "diab_a1c", "diab_med", "ccc_51", "diab_status")
)
head(select(cycle1_diab_data, clinicid, diab_status))
```

**Step 3** — Derive hypertension status. Requires: `cycle1` from Step 1.

```{r, warning=FALSE}
cycle1_htn_data <- recode_after_meds(
  cycle1,
  c(
    # Blood pressure (raw + adjusted)
    "bpmdpbps", "bpmdpbpd", "sbp_adj_mmhg", "dbp_adj_mmhg",
    # Medication inputs (pre-computed in cycle1_with_meds)
    "any_htn_med", "ccc_32",
    # Diabetes chain (input to htn functions)
    "lab_hba1", "diab_a1c", "ccc_51", "diab_med", "diab_status",
    # CVD chain
    "ccc_61", "ccc_63", "ccc_81", "cvd_status",
    # CKD chain
    "lab_bcre", "pgdcgt", "clc_sex", "clc_age", "gfr_ml_min", "ckd_status",
    # Hypertension outcomes
    "htn_status", "htn_adj_status", "htn_control_status", "htn_control_adj_status"
  )
)
head(select(cycle1_htn_data, clinicid, htn_status, htn_adj_status))
```

## Appendix: Installation

```{r, eval=FALSE}
# Install release version from CRAN
install.packages("chmsflow")

# Install the most recent version from GitHub
devtools::install_github("Big-Life-Lab/chmsflow")
```
