---
title: "Recoding CHMS medication variables"
format: html
vignette: >
  %\VignetteIndexEntry{Recoding CHMS medication variables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r results= 'hide', message = FALSE, warning=FALSE}
library(chmsflow)
```

## 1. Introduction

chmsflow provides 16 functions that classify medications from ATC codes recorded in CHMS clinic data. Each function checks whether a respondent is taking a specific drug class and returns `1` (yes) or `0` (no), with `haven::tagged_na()` codes for missing or not-applicable responses.

### Available medication variables

| Variable | Drug class | ATC prefix | Cycles 3-6 function | Cycles 1-2 function |
|----------|------------|------------|----------------------|----------------------|
| `ace_med` | ACE inhibitors | C09 | `is_ace_inhibitor()` | `is_ace_med_cycles1to2()` |
| `bb_med` | Beta blockers | C07 | `is_beta_blocker()` | `is_bb_med_cycles1to2()` |
| `ccb_med` | Calcium channel blockers | C08 | `is_calcium_channel_blocker()` | `is_ccb_med_cycles1to2()` |
| `diur_med` | Diuretics | C03 | `is_diuretic()` | `is_diur_med_cycles1to2()` |
| `misc_htn_med` | Other antihypertensives | mixed | `is_other_antihtn_med()` | `is_misc_htn_med_cycles1to2()` |
| `any_htn_med` | Any antihypertensive | combined | `is_any_antihtn_med()` | `is_any_htn_med_cycles1to2()` |
| `nsaid_med` | NSAIDs | M01A | `is_nsaid()` | `is_nsaid_med_cycles1to2()` |
| `diab_med` | Diabetes medications | A10 | `is_diabetes_med()` | `is_diab_med_cycles1to2()` |

### Cycle differences

Medication data is structured differently across CHMS cycles:

- **Cycles 1-2** store medications in a flat format with up to 80 individual columns (`atc_101a` to `atc_235a` for ATC codes, `mhr_101b` to `mhr_235b` for time last taken). The cycles 1-2 wrapper functions accept all of these columns as parameters.
- **Cycles 3-6** store medications in a multi-row format with two variables per row: `meucatc` (ATC code) and `npi_25b` (time last taken). Each respondent may have multiple rows. After recoding, results must be aggregated by `clinicid`.

## 2. When to use medication recoding

If your analysis requires medication variables, ***always perform medication recoding first***, before recoding any other variables.

Two downstream health outcome variables depend on medication status:

- **Hypertension** - `any_htn_med` is a required input for `derive_hypertension()`, `derive_hypertension_adj()`, `derive_hypertension_control()`, and `derive_hypertension_control_adj()`.
- **Diabetes** - `diab_med` is a required input for `derive_diabetes_status()`.

Because passing all raw medication columns into `rec_with_table()` is cumbersome, chmsflow uses ***intermediate medication variables*** (`any_htn_med2` and `diab_med2`). The workflow is:

1. Recode raw medication variables to create `any_htn_med` and `diab_med`.
2. Duplicate these to create `any_htn_med2` and `diab_med2`.
3. Merge the intermediate variables with the main cycle data.
4. Use the intermediate variables as inputs when recoding chronic condition variables.

This approach eliminates the need to repeatedly reference long lists of raw medication variables.

## 3. Quick start

A minimal example for cycles 3-6: recode ACE inhibitor status, aggregate per respondent, and merge with main cycle data.

```{r, warning=FALSE, eval=FALSE}
library(recodeflow)
library(dplyr)

# Recode ace_med from multi-row medication data
cycle3_ace_data <- rec_with_table(
  cycle3_meds,
  c("clinicid", "meucatc", "npi_25b", "ace_med"),
  variable_details = variable_details
)

# Aggregate to one row per respondent (take max: 1 if any row matched)
cycle3_ace_data <- cycle3_ace_data %>%
  group_by(clinicid) %>%
  summarize(ace_med = max(as.numeric(as.character(ace_med))))

# Merge with main cycle data
cycle3 <- merge(cycle3, cycle3_ace_data, by = "clinicid")
```

## 4. Detailed workflow

### 4.1 Recode medication variables for individual cycles

The medication data object must always be named `cyclex_meds` for the recoding scripts to function correctly. All other data for the same cycle should be stored separately under the name `cyclex`, as specified in `recoding-variables.qmd`.

```{r, warning=FALSE, eval=FALSE}
library(recodeflow)
library(dplyr)

# Recoding basic variables from cycles 2 and 3
cycle2_medication_data <- rec_with_table(
  cycle2_meds,
  c("clinicid", "atc_101a", "mhr_101b"),
  variable_details = variable_details
)
head(cycle2_medication_data)

cycle3_medication_data <- rec_with_table(
  cycle3_meds,
  c("clinicid", "meucatc", "npi_25b"),
  variable_details = variable_details
)
head(cycle3_medication_data)
```

#### Cycles 1-2 (flat format)

For cycles 1-2, all raw ATC code and time-last-taken columns must be listed explicitly in the `rec_with_table()` call:

```{r, warning=FALSE, eval=FALSE}
# Recoding derived variable ace_med for cycle 2
cycle2_ace_medication_data <- rec_with_table(
  cycle2_meds,
  c(
    "clinicid",
    "atc_101a", "atc_102a", "atc_103a", "atc_104a", "atc_105a",
    "atc_106a", "atc_107a", "atc_108a", "atc_109a", "atc_110a",
    "atc_111a", "atc_112a", "atc_113a", "atc_114a", "atc_115a",
    "atc_201a", "atc_202a", "atc_203a", "atc_204a", "atc_205a",
    "atc_206a", "atc_207a", "atc_208a", "atc_209a", "atc_210a",
    "atc_211a", "atc_212a", "atc_213a", "atc_214a", "atc_215a",
    "atc_131a", "atc_132a", "atc_133a", "atc_134a", "atc_135a",
    "atc_231a", "atc_232a", "atc_233a", "atc_234a", "atc_235a",
    "mhr_101b", "mhr_102b", "mhr_103b", "mhr_104b", "mhr_105b",
    "mhr_106b", "mhr_107b", "mhr_108b", "mhr_109b", "mhr_110b",
    "mhr_111b", "mhr_112b", "mhr_113b", "mhr_114b", "mhr_115b",
    "mhr_201b", "mhr_202b", "mhr_203b", "mhr_204b", "mhr_205b",
    "mhr_206b", "mhr_207b", "mhr_208b", "mhr_209b", "mhr_210b",
    "mhr_211b", "mhr_212b", "mhr_213b", "mhr_214b", "mhr_215b",
    "mhr_131b", "mhr_132b", "mhr_133b", "mhr_134b", "mhr_135b",
    "mhr_231b", "mhr_232b", "mhr_233b", "mhr_234b", "mhr_235b",
    "ace_med"
  ),
  variable_details = variable_details
)
head(select(cycle2_ace_medication_data, clinicid, ace_med))
```

#### Cycles 3-6 (multi-row format)

For cycles 3-6, only `meucatc` and `npi_25b` are needed:

```{r, warning=FALSE, eval=FALSE}
# Recoding derived variable ace_med for cycle 3
cycle3_ace_medication_data <- rec_with_table(
  cycle3_meds,
  c("clinicid", "meucatc", "npi_25b", "ace_med"),
  variable_details = variable_details
)
head(cycle3_ace_medication_data)
```

### 4.2 Merge recoded medication data from different cycles

Because cycles 3-6 have multiple rows per respondent, aggregate by `clinicid` before merging:

```{r, warning=FALSE, eval=FALSE}
# Aggregating recoded cycle 3 data by clinicid
cycle3_ace_medication_data <- cycle3_ace_medication_data %>%
  group_by(clinicid) %>%
  summarize(
    meucatc = paste(unique(meucatc), collapse = ", "),
    npi_25b = paste(unique(npi_25b), collapse = ", "),
    ace_med = max(as.numeric(as.character(ace_med)))
  )
head(cycle3_ace_medication_data)

# De-factorizing ace_med variable in recoded cycle 2 data
cycle2_ace_medication_data$ace_med <-
  as.numeric(as.character(cycle2_ace_medication_data$ace_med))

# Merging recoded cycle 2 and cycle 3 data
cycles2and3_ace_medication_data <- merge(
  cycle2_ace_medication_data,
  cycle3_ace_medication_data,
  by = "clinicid"
)

# Keep one ace_med column (1 if either cycle had the medication)
cycles2and3_ace_medication_data <- cycles2and3_ace_medication_data %>%
  mutate(ace_med = pmax(ace_med.x, ace_med.y)) %>%
  select(-c(ace_med.x, ace_med.y))
head(cycles2and3_ace_medication_data)
```

## 5. Integration with health outcomes

### 5.1 Diabetes status

The full pipeline: recode `diab_med` from raw medication data, create the intermediate variable `diab_med2`, merge with main cycle data, then derive diabetes status.

```{r, warning=FALSE, eval=FALSE}
# Step 1: Recode diab_med from medication data
cycle3_diab_medication_data <- rec_with_table(
  cycle3_meds,
  c("clinicid", "meucatc", "npi_25b", "diab_med"),
  variable_details = variable_details
)

# Step 2: Aggregate by clinicid
cycle3_diab_medication_data <- cycle3_diab_medication_data %>%
  group_by(clinicid) %>%
  summarize(
    meucatc = paste(unique(meucatc), collapse = ", "),
    npi_25b = paste(unique(npi_25b), collapse = ", "),
    diab_med = max(as.numeric(as.character(diab_med)))
  )

# Step 3: Create intermediate variable and merge
cycle3_diab_medication_data$diab_med2 <-
  as.numeric(as.character(cycle3_diab_medication_data$diab_med))
cycle3_diab_medication_data <-
  select(cycle3_diab_medication_data, clinicid, diab_med2)

cycle3 <- merge(cycle3, cycle3_diab_medication_data, by = "clinicid")

# Step 4: Derive diabetes status
cycle3_diab_data <- rec_with_table(
  cycle3,
  c("lab_hba1", "diab_a1c", "diab_med2", "ccc_51", "diab_status"),
  variable_details = variable_details
)
head(cycle3_diab_data)
```

### 5.2 Hypertension status

The same pattern applies for hypertension, using `any_htn_med` and the intermediate variable `any_htn_med2`:

```{r, warning=FALSE, eval=FALSE}
# Step 1: Recode any_htn_med from medication data
cycle3_htn_medication_data <- rec_with_table(
  cycle3_meds,
  c("clinicid", "meucatc", "npi_25b", "any_htn_med"),
  variable_details = variable_details
)

# Step 2: Aggregate by clinicid
cycle3_htn_medication_data <- cycle3_htn_medication_data %>%
  group_by(clinicid) %>%
  summarize(
    any_htn_med = max(as.numeric(as.character(any_htn_med)))
  )

# Step 3: Create intermediate variable and merge
cycle3_htn_medication_data$any_htn_med2 <-
  cycle3_htn_medication_data$any_htn_med
cycle3_htn_medication_data <-
  select(cycle3_htn_medication_data, clinicid, any_htn_med2)

cycle3 <- merge(cycle3, cycle3_htn_medication_data, by = "clinicid")

# Step 4: Derive hypertension status
cycle3_htn_data <- rec_with_table(
  cycle3,
  c(
    "bpmdpbps", "bpmdpbpd", "sbp_adj_mmhg", "dbp_adj_mmhg",
    "any_htn_med2", "ccc_32", "cvd_status", "diab_status",
    "ckd_status", "htn_status"
  ),
  variable_details = variable_details
)
head(cycle3_htn_data)
```

## Appendix: Installation

```{r eval=FALSE}
# Install release version from CRAN
install.packages("chmsflow")

# Install the most recent version from GitHub
devtools::install_github("Big-Life-Lab/chmsflow")
```
