---
title: "Recoding CHMS medication variables"
format: html
vignette: >
  %\VignetteIndexEntry{Recoding CHMS medication variables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r results='hide', message=FALSE, warning=FALSE, eval=TRUE}
library(chmsflow)
library(recodeflow)
library(dplyr)
```

## 1. Introduction

chmsflow provides 16 functions that classify medications from ATC codes recorded in CHMS clinic data. Each function checks whether a respondent is taking a specific drug class and returns `1` (yes) or `0` (no), with `haven::tagged_na()` codes for missing or not-applicable responses.

### Available medication variables

| Variable | Drug class | ATC prefix | Cycles 3-6 function | Cycles 1-2 function |
|----------|------------|------------|----------------------|----------------------|
| `ace_med` | ACE inhibitors | C09 | `is_ace_inhibitor()` | `is_ace_med_cycles1to2()` |
| `bb_med` | Beta blockers | C07 | `is_beta_blocker()` | `is_bb_med_cycles1to2()` |
| `ccb_med` | Calcium channel blockers | C08 | `is_calcium_channel_blocker()` | `is_ccb_med_cycles1to2()` |
| `diur_med` | Diuretics | C03 | `is_diuretic()` | `is_diur_med_cycles1to2()` |
| `misc_htn_med` | Other antihypertensives | mixed | `is_other_antihtn_med()` | `is_misc_htn_med_cycles1to2()` |
| `any_htn_med` | Any antihypertensive | combined | `is_any_antihtn_med()` | `is_any_htn_med_cycles1to2()` |
| `nsaid_med` | NSAIDs | M01A | `is_nsaid()` | `is_nsaid_med_cycles1to2()` |
| `diab_med` | Diabetes medications | A10 | `is_diabetes_med()` | `is_diab_med_cycles1to2()` |

### Cycle differences

Medication data is structured differently across CHMS cycles:

- **Cycles 1-2** store medications in a flat format with up to 80 individual columns (`atc_101a` to `atc_235a` for ATC codes, `mhr_101b` to `mhr_235b` for time last taken). The cycles 1-2 wrapper functions accept all of these columns as parameters.
- **Cycles 3-6** store medications in a multi-row format with two variables per row: `meucatc` (ATC code) and `npi_25b` (time last taken). Each respondent may have multiple rows. After recoding, results must be aggregated by `clinicid`.

## 2. When to use medication recoding

If your analysis requires medication variables, ***always perform medication recoding first***, before recoding any other variables. Two downstream health outcome variables depend on medication status:

- **Hypertension** — `any_htn_med` must be merged into the main cycle dataset before deriving hypertension outcomes.
- **Diabetes** — `diab_med` must be merged before deriving diabetes outcomes.

## 3. Workflow

The workflow is the same for all cycles: recode medication variables and merge into the main cycle dataset using `recode_meds_cycles1to2()` or `recode_meds_cycles3to6()`, then derive health outcomes using `recode_after_meds()`. Use `recode_after_meds()` instead of `rec_with_table()` — it automatically excludes medication-specific rows from `variable_details` so pre-computed medication columns are passed through rather than re-derived.

### 3.1 Cycles 1-2

Cycles 1-2 medication data uses uppercase column names (`CLINICID`, `ATC_101A`, etc.). `recode_meds_cycles1to2()` normalizes these internally.

**Step 1** — Recode medication variables and merge with main cycle data. Requires: `cycle1`, `cycle1_meds`.

```{r, warning=FALSE}
cycle1 <- recode_meds_cycles1to2(cycle1, cycle1_meds, c("any_htn_med", "diab_med"))
```

**Step 2** — Derive diabetes status. Requires: `cycle1` from Step 1.

```{r, warning=FALSE}
cycle1_diab_data <- recode_after_meds(
  cycle1,
  c("lab_hba1", "diab_a1c", "diab_med", "ccc_51", "diab_status")
)
head(select(cycle1_diab_data, clinicid, diab_status))
```

**Step 3** — Derive hypertension status. Requires: `cycle1` from Step 1.

```{r, warning=FALSE}
cycle1_htn_data <- recode_after_meds(
  cycle1,
  c(
    # Blood pressure (raw + adjusted)
    "bpmdpbps", "bpmdpbpd", "sbp_adj_mmhg", "dbp_adj_mmhg",
    # Medication inputs (merged in Step 1)
    "any_htn_med", "ccc_32",
    # Diabetes chain (input to htn functions)
    "lab_hba1", "diab_a1c", "ccc_51", "diab_med", "diab_status",
    # CVD chain
    "ccc_61", "ccc_63", "ccc_81", "cvd_status",
    # CKD chain
    "lab_bcre", "pgdcgt", "clc_sex", "clc_age", "gfr_ml_min", "ckd_status",
    # Hypertension outcomes
    "htn_status", "htn_adj_status", "htn_control_status", "htn_control_adj_status"
  )
)
head(select(cycle1_htn_data, clinicid, htn_status, htn_adj_status))
```

### 3.2 Cycles 3-6

**Step 1** — Recode medication variables and merge with main cycle data. Requires: `cycle3`, `cycle3_meds`.

```{r, warning=FALSE}
cycle3 <- recode_meds_cycles3to6(cycle3, cycle3_meds, c("any_htn_med", "diab_med"))
```

**Step 2** — Derive diabetes status. Requires: `cycle3` from Step 1.

```{r, warning=FALSE}
cycle3_diab_data <- recode_after_meds(
  cycle3,
  c("lab_hba1", "diab_a1c", "diab_med", "ccc_51", "diab_status")
)
head(select(cycle3_diab_data, clinicid, diab_status))
```

**Step 3** — Derive hypertension status. Requires: `cycle3` from Step 1.

`cvd_status`, `diab_status`, and `ckd_status` are intermediate inputs to the hypertension functions. Their full input chains must also be listed so `recode_after_meds()` can derive them.

```{r, warning=FALSE}
cycle3_htn_data <- recode_after_meds(
  cycle3,
  c(
    # Blood pressure (raw + adjusted)
    "bpmdpbps", "bpmdpbpd", "sbp_adj_mmhg", "dbp_adj_mmhg",
    # Medication inputs (merged in Step 1)
    "any_htn_med", "ccc_32",
    # Diabetes chain (input to htn functions)
    "lab_hba1", "diab_a1c", "ccc_51", "diab_med", "diab_status",
    # CVD chain
    "ccc_61", "ccc_63", "ccc_81", "cvd_status",
    # CKD chain
    "lab_bcre", "pgdcgt", "clc_sex", "clc_age", "gfr_ml_min", "ckd_status",
    # Hypertension outcomes
    "htn_status", "htn_adj_status", "htn_control_status", "htn_control_adj_status"
  )
)
head(select(cycle3_htn_data, clinicid, htn_status, htn_adj_status))
```

## 4. Advanced: using individual classification functions

The `is_*` functions underlie the wrapper functions and are available directly for custom workflows — for example, deriving a single drug class without the full pipeline, or integrating classification logic into your own aggregation steps.

Each function accepts an ATC code and a time-last-taken value and returns `1`, `0`, or a `tagged_na()` code:

```{r, warning=FALSE}
# Single medication classification
is_beta_blocker("C07AA05", 1) # returns 1
is_ace_inhibitor("C09AA02", 1) # returns 1
is_diabetes_med("A10BA02", 1) # returns 1
```

### Cycle format differences

**Cycles 1-2** — one row per respondent with up to 80 `atc_*/mhr_*` column pairs. The `is_*_med_cycles1to2()` variants accept named arguments for each slot:

```{r, warning=FALSE}
# Classification using cycles 1-2 wide-format columns
is_ace_med_cycles1to2(atc_101a = "C09AA02", mhr_101b = 1) # returns 1
is_ace_med_cycles1to2(atc_101a = "C09AA02", mhr_101b = 6) # returns 0 (not taken recently)
```

**Cycles 3-6** — one row per medication per respondent with two columns: `meucatc` (ATC code) and `npi_25b` (time last taken). Classify per row, then aggregate across rows per respondent:

```{r, warning=FALSE}
cycle3_meds |>
  mutate(ace_med = is_ace_inhibitor(meucatc, npi_25b)) |>
  group_by(clinicid) |>
  summarize(ace_med = max(as.numeric(as.character(ace_med))))
```

## Appendix: Installation

```{r, eval=FALSE}
# Install release version from CRAN
install.packages("chmsflow")

# Install the most recent version from GitHub
devtools::install_github("Big-Life-Lab/chmsflow")
```
