---
title: "Recoding CHMS medication variables"
format: html
vignette: >
  %\VignetteIndexEntry{Recoding CHMS medication variables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## 1. Load packages

```{r eval= FALSE}
# Install release version from CRAN
install.packages("chmsflow")

# Install the most recent version from GitHub
devtools::install_github("Big-Life-Lab/chmsflow")
```

```{r results= 'hide', message = FALSE, warning=FALSE}
library(chmsflow)
```

## 2. Recode medication variables for individual cycles

The medication data object must always be named cyclex_meds for the recoding scripts to function correctly. All other data for the same cycle should be stored separately under the name cyclex, as specified in `recoding-variables.qmd`.

```{r, warning=FALSE, eval=FALSE}
# Load recodeflow and dplyr
library(recodeflow)
library(dplyr)

# Recoding basic variables from cycles 1 and 3
cycle2_medication_data <- rec_with_table(cycle2_meds, c("clinicid", "atc_101a", "mhr_101b"), variable_details = variable_details)
head(cycle2_medication_data)

cycle3_medication_data <- rec_with_table(cycle3_meds, c("clinicid", "meucatc", "npi_25b"), variable_details = variable_details)
head(cycle3_medication_data)

# Recoding derived variable ace_med for cycle2.
cycle2_ace_medication_data <- rec_with_table(cycle2_meds, c("clinicid", "atc_101a", "atc_102a", "atc_103a", "atc_104a", "atc_105a", "atc_106a", "atc_107a", "atc_108a", "atc_109a", "atc_110a", "atc_111a", "atc_112a", "atc_113a", "atc_114a", "atc_115a", "atc_201a", "atc_202a", "atc_203a", "atc_204a", "atc_205a", "atc_206a", "atc_207a", "atc_208a", "atc_209a", "atc_210a", "atc_211a", "atc_212a", "atc_213a", "atc_214a", "atc_215a", "atc_131a", "atc_132a", "atc_133a", "atc_134a", "atc_135a", "atc_231a", "atc_232a", "atc_233a", "atc_234a", "atc_235a", "mhr_101b", "mhr_102b", "mhr_103b", "mhr_104b", "mhr_105b", "mhr_106b", "mhr_107b", "mhr_108b", "mhr_109b", "mhr_110b", "mhr_111b", "mhr_112b", "mhr_113b", "mhr_114b", "mhr_115b", "mhr_201b", "mhr_202b", "mhr_203b", "mhr_204b", "mhr_205b", "mhr_206b", "mhr_207b", "mhr_208b", "mhr_209b", "mhr_210b", "mhr_211b", "mhr_212b", "mhr_213b", "mhr_214b", "mhr_215b", "mhr_131b", "mhr_132b", "mhr_133b", "mhr_134b", "mhr_135b", "mhr_231b", "mhr_232b", "mhr_233b", "mhr_234b", "mhr_235b", "ace_med"), variable_details = variable_details)
head(select(cycle2_ace_medication_data, clinicid, ace_med))

# Recoding derived variable ace_med for cycle3. Much simpler function used here.
cycle3_ace_medication_data <- rec_with_table(cycle3_meds, c("clinicid", "meucatc", "npi_25b", "ace_med"), variable_details = variable_details)
head(cycle3_ace_medication_data)
```

## 3. Merge recoded medication data from different cycles

```{r, warning=FALSE, eval=FALSE}
# Aggregating recoded cycle3 data by clinicid
cycle3_ace_medication_data <- cycle3_ace_medication_data %>%
  group_by(clinicid) %>%
  summarize(
    meucatc = paste(unique(meucatc), collapse = ", "), # Concatenate unique values of meucatc
    npi_25b = paste(unique(npi_25b), collapse = ", "), # Concatenate unique values of npi_25b
    ace_med = max(as.numeric(as.character(ace_med))) # Find maximum of ace_med
  )
head(cycle3_ace_medication_data)

# De-factorizing ace_med variable in recoded cycle2 data
cycle2_ace_medication_data$ace_med <- as.numeric(as.character(cycle2_ace_medication_data$ace_med))

# Merging recoded cycle2 and cycle3 data, and ensuring one ace_med column in the end
cycles2and3_ace_medication_data <- merge(cycle2_ace_medication_data, cycle3_ace_medication_data, by = "clinicid")

cycles2and3_ace_medication_data <- cycles2and3_ace_medication_data %>%
  mutate(ace_med = pmax(ace_med.x, ace_med.y)) %>%
  select(-c(ace_med.x, ace_med.y))
head(cycles2and3_ace_medication_data)
```

## 4. Example: Determine diabetes status by recoding medications

If your analysis requires medication variables, ***always perform medication recoding first***, before recoding any other variables.

Derived medication variables may be needed to derive certain chronic condition variables, such as hypertension (needing `any_htn_med`) and diabetes (needing `diab_med`). However, passing all required raw medication variables into a single `rec_with_table()` call is cumbersome. For cycles 1-2, this means listing many variables (`atc_101a`, `atc_102a`, etc.), while cycles 3-6 involve medication variables with multiple rows per respondent (`meucatc` and `npi_25b`).

To simplify this process, use ***intermediate medication variables*** `any_htn_med2` and `diab_med2`. The workflow is: (1) recode raw medication variables to create `any_htn_med` and `diab_med`, (2) duplicate these to create `any_htn_med2` and `diab_med2`, (3) merge the intermediate variables with the main cycle data, and (4) use the intermediate variables as inputs when recoding chronic condition variables. This approach eliminates the need to repeatedly reference long lists of raw medication variables.

```{r, warning=FALSE, eval=FALSE}
# Recode raw and derived medication variables first, including diab_med for diabetes
cycle3_diab_medication_data <- rec_with_table(cycle3_meds, c("clinicid", "meucatc", "npi_25b", "diab_med"), variable_details = variable_details)

# Aggregating recoded cycle3 medication data by clinicid
cycle3_diab_medication_data <- cycle3_diab_medication_data %>%
  group_by(clinicid) %>%
  summarize(
    meucatc = paste(unique(meucatc), collapse = ", "), # Concatenate unique values of meucatc
    npi_25b = paste(unique(npi_25b), collapse = ", "), # Concatenate unique values of npi_25b
    diab_med = max(as.numeric(as.character(diab_med))) # Find maximum of diab_med
  )

# Create intermediate medication variable by duplicating recoded derived medication variable
cycle3_diab_medication_data$diab_med2 <- as.numeric(as.character(cycle3_diab_medication_data$diab_med))
cycle3_diab_medication_data <- select(cycle3_diab_medication_data, clinicid, diab_med2)

# Merge medication data with original data
# Call overall data "cyclex" in order to recode other variables afterwards
cycle3 <- merge(cycle3, cycle3_diab_medication_data, by = "clinicid")

# Recode diabetes status using recoded, intermediate medication variables
cycle3_diab_data <- rec_with_table(cycle3, c("lab_hba1", "diab_a1c", "diab_med2", "ccc_51", "diab_status"), variable_details = variable_details)
head(cycle3_diab_data)
```
