---
title: "Recoding CHMS medication variables"
format: html
---

1. Load required packages, functions, and worksheets (variables and variable details sheets). See README.md in GitHub to see how packages can be installed at the RDC. SECTION WILL BE REMOVED WHEN PACKAGE POSTED!
```{r}
library(here) # Ensure here is set to your working directory
library(dplyr)
library(recodeflow)

source(here("R", "medications.R"), local = knitr::knit_global())
source(here("R", "family-history.R"), local = knitr::knit_global())
source(here("R", "diabetes.R"), local = knitr::knit_global())
source(here("R", "kidney.R"), local = knitr::knit_global())
source(here("R", "blood-pressure.R"), local = knitr::knit_global())

variables <- read.csv(here("inst", "extdata", "variables.csv"))
variable_details <- read.csv(here("inst", "extdata", "variable-details.csv"))
```

2. Create synthetic data for cycle 2 and cycle 3, with each cycle's medication data being separate. Medication data object always has to be called "cyclex_meds" in order for recoding to work, with rest of the data being called "cyclex" as stated in recoding-variables.qmd. Note that row headers for medication data of cycles 1, 4, and 6 must be put to lower case prior to recoding.
```{r}
# cycle 2 - wide medication data
cycle2 <- data.frame(
  clinicid = c(1, 2, 3, 4),
  clc_age = c(20, 30, 40, 50),
  clc_sex = c(1, 2, 1, 2),
  sdcdcgt = c(1, 1, 1, 2),
  lab_hba1 = c(0.07, 0.06, 0.05, 0.04),
  lab_bcre = c(20, 30, 40, 50),
  bpmdpbps = c(140, 130, 120, 110),
  bpmdpbpd = c(95, 85, 75, 65),
  ccc_51 = c(1, 2, 2, 2),
  ccc_61 = c(2, 2, 2, 2),
  ccc_63 = c(2, 2, 2, 2),
  ccc_81 = c(1, 2, 2, 1),
  ccc_32 = c(2, 1, 1, 1)
)

cycle2_meds <- data.frame(
  clinicid = c(1, 2, 3, 4),
  atc_101a = c("C07AA05", "C09AA06", "C08CA01", "A10BC02"),
  atc_102a = c("A01AB05", "C09AA02", "C07AB02", "C03AA03"),
  atc_103a = c("C02CC07", "C08CA06", "C07AB07", "C07AB03"),
  atc_104a = c("C03CA01", "C03CA01", "C09AA04", "C08CA02"),
  atc_105a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_106a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_107a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_108a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_109a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_110a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_111a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_112a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_113a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_114a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_115a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_201a = c("C03BA08", "C07AA07", "C07AA12", "M01AG02"),
  atc_202a = c("C07AA05", "C09AA06", "C08CA01", "A10BC02"),
  atc_203a = c("C03BA08", "C07AA07", "C07AA12", "M01AG02"),
  atc_204a = c("C03BC02", "C08CA01", "C07AA05", "C07AA06"),
  atc_205a = c("C02KX01", "C02AA05", "C07AG02", "C07AA06"),
  atc_206a = c("A01AB05", "C09AA02", "C07AB02", "C03AA03"),
  atc_207a = c("C02CC07", "C08CA06", "C07AB07", "C07AB03"),
  atc_208a = c("C03CA01", "C03CA01", "C09AA04", "C08CA02"),
  atc_209a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_210a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_211a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_212a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_213a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_214a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_215a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_131a = c("C07AA05", "C09AA06", "C08CA01", "A10BC02"),
  atc_132a = c("C03BA08", "C07AA07", "C07AA12", "M01AG02"),
  atc_133a = c("C03BC02", "C08CA01", "C07AA05", "C07AA06"),
  atc_134a = c("C02KX01", "C02AA05", "C07AG02", "C07AA06"),
  atc_135a = c("A01AB05", "C09AA02", "C07AB02", "C03AA03"),
  atc_231a = c("C02CC07", "C08CA06", "C07AB07", "C07AB03"),
  atc_232a = c("C03CA01", "C03CA01", "C09AA04", "C08CA02"),
  atc_233a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_234a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  atc_235a = c("C09AA04", "C07AB02", "C08CA02", "A10BD05"),
  mhr_101b = c(2, 3, 1, 3),
  mhr_102b = c(1, 4, 3, 2),
  mhr_103b = c(3, 2, 1, 4),
  mhr_104b = c(1, 2, 4, 3),
  mhr_105b = c(3, 2, 1, 4),
  mhr_106b = c(3, 2, 1, 4),
  mhr_107b = c(3, 2, 1, 4),
  mhr_108b = c(3, 2, 1, 4),
  mhr_109b = c(3, 2, 1, 4),
  mhr_110b = c(3, 2, 1, 4),
  mhr_111b = c(3, 2, 1, 4),
  mhr_112b = c(3, 2, 1, 4),
  mhr_113b = c(3, 2, 1, 4),
  mhr_114b = c(3, 2, 1, 4),
  mhr_115b = c(3, 2, 1, 4),
  mhr_201b = c(3, 2, 2, 3),
  mhr_202b = c(2, 3, 1, 3),
  mhr_203b = c(3, 2, 2, 3),
  mhr_204b = c(3, 1, 5, 1),
  mhr_205b = c(1, 2, 3, 4),
  mhr_206b = c(1, 4, 3, 2),
  mhr_207b = c(3, 2, 1, 4),
  mhr_208b = c(1, 2, 4, 3),
  mhr_209b = c(3, 2, 1, 4),
  mhr_210b = c(3, 2, 1, 4),
  mhr_211b = c(3, 2, 1, 4),
  mhr_212b = c(3, 2, 1, 4),
  mhr_213b = c(3, 2, 1, 4),
  mhr_214b = c(3, 2, 1, 4),
  mhr_215b = c(3, 2, 1, 4),
  mhr_131b = c(2, 3, 1, 3),
  mhr_132b = c(3, 2, 2, 3),
  mhr_133b = c(3, 1, 5, 1),
  mhr_134b = c(1, 2, 3, 4),
  mhr_135b = c(1, 4, 3, 2),
  mhr_231b = c(3, 2, 1, 4),
  mhr_232b = c(1, 2, 4, 3),
  mhr_233b = c(3, 2, 1, 4),
  mhr_234b = c(3, 2, 1, 4),
  mhr_235b = c(3, 2, 1, 4)
) 

# Cycle 3 - long medication data
cycle3_survey <- data.frame(
  clinicid = c(1, 2, 3, 4),
  clc_age = c(20, 30, 40, 50),
  clc_sex = c(1, 2, 1, 2),
  pgdcgt = c(1, 1, 1, 2),
  lab_hba1 = c(0.07, 0.06, 0.05, 0.04),
  lab_bcre = c(20, 30, 40, 50),
  bpmdpbps = c(140, 130, 120, 110),
  bpmdpbpd = c(95, 85, 75, 65),
  ccc_51 = c(1, 2, 2, 2),
  ccc_61 = c(2, 2, 2, 2),
  ccc_63 = c(2, 2, 2, 2),
  ccc_81 = c(1, 2, 2, 1),
  ccc_32 = c(2, 1, 1, 1)
) 

num_vars_cycle3 <- ncol(cycle3) - 1
clinicids <- unique(cycle3$clinicid)
n_clinicids <- length(clinicids)
  
cycle3_meds <- data.frame(
  clinicid = rep(clinicids, each = num_vars_cycle3),
  meucatc = rep(c("C07AA05", "C09AA06", "C08CA01", "A10BC02"), length.out = num_vars_cycle3 * n_clinicids),
  npi_25b = rep(c(2, 3, 1, 3), length.out = num_vars_cycle3 * n_clinicids)
) 
```

3. Recode medication variables for individual cycles.
```{r}
# Recoding basic variables from cycles 1 and 3
cycle2_medication_data <- rec_with_table(cycle2_meds, c("clinicid", "atc_101a", "mhr_101b"), variable_details = variable_details)
head(cycle2_medication_data)

cycle3_medication_data <- rec_with_table(cycle3_meds, c("clinicid", "meucatc", "npi_25b"), variable_details = variable_details)
head(cycle3_medication_data)

# Recoding derived variable acemed for cycle2. Select vars by role used to avoid writing all med variables in one line
cycle2_ace_medication_data <- rec_with_table(cycle2_meds, c("clinicid", "acemed", recodeflow:::select_vars_by_role("Drugs", variables)), variable_details = variable_details)
head(select(cycle2_ace_medication_data, clinicid, acemed)) 

# Recoding derived variable acemed for cycle3. Much simpler function used here.
cycle3_ace_medication_data <- rec_with_table(cycle3_meds, c("clinicid", "meucatc", "npi_25b", "acemed"), variable_details = variable_details)
head(cycle3_ace_medication_data)
```

4. Merge recoded medication data from different cycles.
```{r}
# Aggregating recoded cycle3 data by clinicid
cycle3_ace_medication_data <- cycle3_ace_medication_data %>%
  group_by(clinicid) %>%
  summarize(
    meucatc = paste(unique(meucatc), collapse = ", "),  # Concatenate unique values of meucatc
    npi_25b = paste(unique(npi_25b), collapse = ", "),  # Concatenate unique values of npi_25b
    acemed = max(as.numeric(as.character(acemed))) # Find maximum of acemed
  )
head(cycle3_ace_medication_data)

# De-factorizing acemed variable in recoded cycle2 data
cycle2_ace_medication_data$acemed <- as.numeric(as.character(cycle2_ace_medication_data$acemed))

# Merging recoded cycle2 and cycle3 data, and ensuring one acemed column in the end
cycles2and3_ace_medication_data <- merge(cycle2_ace_medication_data, cycle3_ace_medication_data, by = "clinicid")

cycles2and3_ace_medication_data <- cycles2and3_ace_medication_data %>%
  mutate(acemed = pmax(acemed.x, acemed.y)) %>%
  select(-c(acemed.x, acemed.y))
head(cycles2and3_ace_medication_data)
```

5. If you need medication variables, always recode them before other variables.
Example: Determine hypertension status by recoding medications 
```{r}
# Cycles 1-2
# Recode medication variables first - anymed for hypertension and diab_drug for diabetes which is involved in determining hypertension status
cycle2_htn_medication_data <- rec_with_table(cycle2_meds, c("clinicid", recodeflow:::select_vars_by_role("Drugs", variables)), variable_details = variable_details)

# Create dummy medication variable by duplicating recoded medication variable
cycle2_htn_medication_data$anymed2 <- as.numeric(as.character(cycle2_htn_medication_data$anymed))
cycle2_htn_medication_data$diab_drug2 <- as.numeric(as.character(cycle2_htn_medication_data$diab_drug))
cycle2_htn_medication_data <- select(cycle2_htn_medication_data, clinicid, anymed2, diab_drug2)

# Merge medication data with rest of data
cycle2 <- merge(cycle2, cycle2_htn_medication_data, by = "clinicid")

# Recode hypertension (and diabetes) status using recoded, dummy medication variables
cycle2_htn_data <- rec_with_table(cycle2, c(recodeflow:::select_vars_by_role("Test", variables)), variable_details = variable_details)
head(cycle2_htn_data)

# Cycles 3-6
# Recode medication variables first - - anymed for hypertension and diab_drug for diabetes which is involved in determining hypertension status
cycle3_htn_medication_data <- rec_with_table(cycle3_meds, c("clinicid", "meucatc", "npi_25b", "anymed", "diab_drug"), variable_details = variable_details)

# Aggregating recoded cycle3 data by clinicid
cycle3_htn_medication_data <- cycle3_htn_medication_data %>%
  group_by(clinicid) %>%
  summarize(
    meucatc = paste(unique(meucatc), collapse = ", "),   # Concatenate unique values of meucatc
    npi_25b = paste(unique(npi_25b), collapse = ", "),   # Concatenate unique values of npi_25b
    anymed = max(as.numeric(as.character(anymed))),      # Find maximum of anymed
    diab_drug = max(as.numeric(as.character(diab_drug))) # Find maximum of diab_drug
  )

# Create dummy medication variable by duplicating recoded medication variable
cycle3_htn_medication_data$anymed2 <- as.numeric(as.character(cycle3_htn_medication_data$anymed))
cycle3_htn_medication_data$diab_drug2 <- as.numeric(as.character(cycle3_htn_medication_data$diab_drug))
cycle3_htn_medication_data <- select(cycle3_htn_medication_data, clinicid, anymed2, diab_drug2)

# Merge medication data with original data
# Call overall data "cyclex" in order to recode other variables afterwards
cycle3 <- merge(cycle3, cycle3_htn_medication_data, by = "clinicid")

# Recode hypertension (and diabetes) status using recoded, dummy medication variables
cycle3_htn_data <- rec_with_table(cycle3, c(recodeflow:::select_vars_by_role("Test", variables)), variable_details = variable_details)
head(cycle3_htn_data)
```