---
title: "Using chmsflow at an RDC"
format: html
vignette: >
  %\VignetteIndexEntry{Using chmsflow at an RDC}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

Statistics Canada Research Data Centres (RDCs) provide access to CHMS master files in a controlled environment with no internet access. This means you cannot install packages from CRAN or GitHub inside the RDC. This vignette explains how to set up chmsflow at an RDC and how to organize your workflow between the RDC and your regular development environment.

## Workflow

Working with chmsflow at an RDC follows a three-step model:

1. **Develop outside the RDC** -- Write and test your R scripts using chmsflow's bundled dummy data on your own computer. The dummy data mirrors the structure of the real CHMS data, so your code will run the same way inside the RDC.
2. **Have the RDC analyst set up the environment** -- The RDC analyst places the R packages and CHMS data files into the RDC environment (typically on your P drive).
3. **Run inside the RDC** -- Go into the RDC and run your tested scripts against the real CHMS master file data.

This separation means you can iterate on your analysis code without using RDC time, and only go in when the code is ready to run on real data.

## Setup inside the RDC

### 1. Set the library path

Since there is no internet access at the RDC, all R package dependencies (chmsflow, recodeflow, dplyr, haven, purrr, etc.) must be pre-installed in a folder on your P drive. The RDC analyst can do this for you. Then point R to that folder:

```{r, eval=FALSE}
.libPaths("<path_to_your_P_drive_package_folder>")
```

Note that `renv` does not work at the RDC because `renv::restore()` requires internet access to download packages.

### 2. Load packages

```{r, eval=FALSE}
library(chmsflow)
library(recodeflow)
library(dplyr)
```

### 3. Load additional scripts and worksheets

If you have custom functions or modified metadata files, load them with `source()` and `read.csv()`:

```{r, eval=FALSE}
source("my_custom_functions.R")
my_variable_details <- read.csv("my_variable_details.csv")
```

### 4. Load CHMS data

Load CHMS data files and name the objects to match chmsflow's naming conventions. This is important because `rec_with_table()` uses the object name to match the `databaseStart` column in `variable-details.csv`.

- Name main cycle data as `cycle1` through `cycle6`
- Name medication data as `cycle1_meds` through `cycle6_meds`

```{r, eval=FALSE}
# Cycles 3-6: Stata format
cycle3 <- haven::read_stata("cycle3.dta")
cycle4 <- haven::read_stata("cycle4.dta")
cycle5 <- haven::read_stata("cycle5.dta")
cycle6 <- haven::read_stata("cycle6.dta")

# Medication data: Stata for cycles 3-6, SAS for cycles 1-2
cycle3_meds <- haven::read_stata("cycle3_meds.dta")
cycle4_meds <- haven::read_stata("cycle4_meds.dta")
cycle5_meds <- haven::read_stata("cycle5_meds.dta")
cycle6_meds <- haven::read_stata("cycle6_meds.dta")

cycle1_meds <- haven::read_sas("cycle1_meds.sas7bdat")
cycle2_meds <- haven::read_sas("cycle2_meds.sas7bdat")
```

Some cycles use uppercase column names. Lowercase them so `rec_with_table()` can match the variable names in `variable-details.csv`:

```{r, eval=FALSE}
names(cycle1) <- tolower(names(cycle1))
names(cycle4) <- tolower(names(cycle4))
names(cycle6) <- tolower(names(cycle6))

names(cycle1_meds) <- tolower(names(cycle1_meds))
names(cycle4_meds) <- tolower(names(cycle4_meds))
names(cycle6_meds) <- tolower(names(cycle6_meds))
```

### 5. Run your analysis

From here, the workflow is identical to what you tested with dummy data. See `vignette("analysis_walkthrough")` for a complete example, or `vignette("recoding_medications")` for the medication recoding pipeline.

## Tips

- **Test thoroughly outside the RDC.** RDC time is limited. Use chmsflow's dummy data to verify your code runs end-to-end before going in.
- **Bring all dependencies.** Check what chmsflow imports by running `tools::package_dependencies("chmsflow", recursive = TRUE)` outside the RDC, and make sure all of those packages are on your P drive.
- **Keep medication data separate.** Do not merge medication files into the main cycle files before loading. chmsflow expects them as separate objects for `recode_meds_cycles1to2()` and `recode_meds_cycles3to6()`.
